Forcing for Isabelle

### 2018-12-02 LICS/forcing.tex
mejoro un poco y pongo convenciones de notación para variables

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Forcing}
Given a ctm $M$, and an $M$-generic filter $G\sbq\PP$, the Forcing
Theorems relate satisfaction of a formula 
$\phi$ in the generic extension $M[G]$ to the satisfaction of another
formula $\phi'$ in $M$. The map $\phi\mapsto\phi'$ is defined by
recursion on 
the structure of $\phi$. The fact that this map works (in particular,
in order to show that $M[G]\models\ZFC$), we must use the fact that
$M\models\ZFC$. Since we can only assert the latter using internalized
formulas, the aformentioned map will be defined as a function from the
set \formula{} into itself.

@@

We will now make more precise the definition of the map
$\phi\mapsto\phi'$ and how it relates satisfaction in $M$ to that in
$M[G]$. Actually, if the formula $\phi$ has $n$ free variables,
$\phi'$ will have $n+4$ free variables, where the first four account
for the forcing notion and a particular condition. If
$\phi=\phi(x_1,\dots,x_n)$, the standard notation for $\phi'$ is
\[
p\forces_{\PP,\leq,\1}^* \phi(t_1,\dots,t_n).
\]
Here, $\PP,\leq,\1,$ and $p\in\PP$ are the extra parameters. 
%% For the
%% time being, we just consider $\tau_1,\dots,\tau_n$ to be ordinary
%% variables. 
Given $\quine\phi\in\formula$, we will write this  in
Isabelle/ZF by using  
\[
\forceisa(\quine\phi,\PP,\leq,\1,p)
\]
Afterwards, the \emph{forcing relation} $\forces$ is defined by
relativizing/interpreting $\forces^*$ in a ctm $M$, for fixed
$\lb\PP,\leq,\1\rb\in M$:
\[
p\forces \phi(t_1,\dots,t_n) \defi 
\bigl(p\forces_{\PP,\leq,\1}^* \phi(t_1,\dots,t_n)\bigr)^M\!.
\]

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Forcing}
Given a ctm $M$, and an $M$-generic filter $G\sbq\PP$, the Forcing
Theorems relate satisfaction of a formula 
$\phi$ in the generic extension $M[G]$ to the satisfaction of another
formula $\forceisa(\phi)$ in $M$. The map $\forceisa$ is defined by
recursion on 
the structure of $\phi$. The fact that this map works (in particular,
in order to show that $M[G]\models\ZFC$), we must use the fact that
$M\models\ZFC$. Since we can only assert the latter using internalized
formulas, the aformentioned map will be defined as a function from the
set \formula{} into itself.

@@

We will now make more precise the definition of the map
$\forceisa$ and how it relates satisfaction in $M$ to that in
$M[G]$. Actually, if the formula $\phi$ has $n$ free variables,
$\forceisa(\phi)$ will have $n+4$ free variables, where the first four account
for the forcing notion and a particular condition.

In referring to formulas with free variables and their interpretation,
we use the following conventions. If  $M$ is a set, $a_0,\dots,a_n$
are elements of $M$, and $\phi$ has its free variables
among $x_0,\dots,x_n$, we write
\[
M,[a_0,\dots,a_n] \models \phi(x_0,\dots,x_n)
\]
to denote that $\phi$ is satisfied by $M$ when $x_i$ is interpreted
to be $a_i$ ($i=0,\dots,n$). We call the list $[a_0,\dots,a_n]$ the
\emph{environment}. In the case of a formula of the form
$\forceisa(\phi)$, we'll make an abuse of notation and indicate the
variables inside the argument of $\forceisa$. As an example, take the
formula $\phi\defi x_1\in x_0$. Then
\[
M,[a,b] \models x_1\in x_0
\]
will hold whenever $b\in a$; and instead of writing $\forceisa(\phi)$
we will write $\forceisa(x_5\in x_4)$, as in
\[
M,[\PP,\leq,\1,p,\tau,\rho] \models \forceisa(x_5\in x_4).
\]

If
$\phi=\phi(x_0,\dots,x_n)$, the standard notation for $\forceisa(\phi)$ is
\[
p\forces_{\PP,\leq,\1}^* \phi(x_0,\dots,x_n).
\]
Here, the extra parameters are $\PP,\leq,\1,$ and $p\in\PP$, and the
first three are usually omitted. %(since the forcing notion is fixed
% throughout the discussion).  
Afterwards, the \emph{forcing relation}
$\forces$ is defined by 
relativizing/interpreting $\forces^*$ in a ctm $M$, for fixed
$\lb\PP,\leq,\1\rb\in M$: $p\forces \phi(\tau_0,\dots,\tau_n)$ holds
if and only if
\[ 
M,[\PP,\leq,\1,p,\tau_0,\dots,\tau_n]\models \forceisa(x_4,\dots,x_{n+4}).
\]
